---
title: "Lecture 3: Introduction (continued)"
author: "Mingze Huang"
date: "`r Sys.Date()`"
output: slidy_presentation
---

## Chain rule in CPI (Not Chained CPI)


```{r, echo = FALSE, results = 'asis'}
table1 = data.frame(matrix(c(2019, 2020, 2021, 0.02, 0.035, 0.013, 25, 20, 30, 0.25, 0.2, 0.2, 2, 1.5, 3), 3, 5)); colnames(table1) = c("Year", "Weight (Cars)", "Price (cars)", "Weight (Gas)", "Price (Gas)")
knitr::kable(table1, caption = "Illustrate CPI calculation")
``` 
Take 2020 as base year, assume a representative (average) consumer spend 70% of money on cars, 30% of money on gasoline on 2020. By construction we have Cost of (base year) market basket for 2019~2021 is:

$$
C_{2019}=P_{2019}^{C}W_{2020}^{C}+P_{2019}^{G}W_{2020}^{G}=25\times0.035+2\times0.2=1.275\\
C_{2020}=P_{2020}^{C}W_{2020}^{C}+P_{2020}^{G}W_{2020}^{G}=20\times0.035+1.5\times0.2=1\\
C_{2021}=P_{2021}^{C}W_{2020}^{C}+P_{2021}^{G}W_{2020}^{G}=30\times0.035+3\times0.2=1.65
$$


## Chain rule in CPI

By definition, CPI for 2019~2021 is: 
$$
CPI_{2019}=\frac{C_{2019}}{C_{2020}}\times100=127.5\\
CPI_{2020}=\frac{C_{2020}}{C_{2020}}\times100=100\\
CPI_{2021}=\frac{C_{2021}}{C_{2020}}\times100=165
$$
$\$1650$'s purchase power in 2021, probably equivalent to $\$1650\times\frac{127.5}{165}=\$1270$ in 2019!

CPI inflation for 2020~2021 is:

$$
\pi_{2020}=\frac{CPI_{2020}-CPI_{2019}}{CPI_{2019}}\times100\%=\frac{100-127.5}{127.5}\times100\%=-21.57\%\\
\pi_{2021}=\frac{CPI_{2021}-CPI_{2020}}{CPI_{2020}}\times100\%=65\%
$$

## Chain rule in CPI

[In practice, BLS use market basket 24 months ago to calculate CPI.](https://www.bls.gov/cpi/additional-resources/chained-cpi-questions-and-answers.htm)

```{r, echo = FALSE, results = 'asis'}
table1 = data.frame(matrix(c(2019, 2020, 2021, 0.02, 0.035, 0.013, 25, 20, 30, 0.25, 0.2, 0.2, 2, 1.5, 3), 3, 5)); colnames(table1) = c("Year", "Weight (Cars)", "Price (cars)", "Weight (Gas)", "Price (Gas)")
knitr::kable(table1, caption = "Illustrate CPI calculation")
``` 
Alternatively take 2019 as base year. Note that we assume a representative consumer spend half of money on cars and half on gasoline on 2019. By construction we have Cost of (base year) market basket for 2019~2021 is:
$$
C_{2019}=P_{2019}^{C}W_{2019}^{C}+P_{2019}^{G}W_{2019}^{G}=25\times0.02+2\times0.25=1\\
C_{2020}=P_{2020}^{C}W_{2019}^{C}+P_{2020}^{G}W_{2019}^{G}=20\times0.02+1.5\times0.25=0.775\\
C_{2021}=P_{2021}^{C}W_{2019}^{C}+P_{2021}^{G}W_{2019}^{G}=30\times0.02+3\times0.25=1.35
$$

## Chain rule in CPI

CPI for 2019~2021 is:
$$
CPI_{2019}=\frac{C_{2019}}{C_{2019}}\times100=100\\
CPI_{2020}=\frac{C_{2020}}{C_{2019}}\times100=77.5\\
CPI_{2021}=\frac{C_{2021}}{C_{2020}}\times100=135
$$


$\$1650$'s purchase power in 2021, probably equivalent to $\$1650\times\frac{100}{135}=\$1222.22$ in 2019!

Base year matters because your market basket changes!

CPI inflation for 2020~2021 is:

$$
\pi_{2020}=\frac{CPI_{2020}-CPI_{2019}}{CPI_{2019}}\times100\%=\frac{77.5-100}{100}=-22.5\%\\
\pi_{2021}=\frac{CPI_{2021}-CPI_{2020}}{CPI_{2020}}\times100\%=\frac{135-77.5}{77.5}=74.19\%
$$

[CPI vs. Chained CPI](https://www.bls.gov/cpi/additional-resources/chained-cpi-questions-and-answers.htm)
